---
title: "Cardiovascular Disease Data Analysis"
author: Giovanni Billo, Muhammad Mubashar Shahzad, Carlos Velázquez Fernández, Fabio Vicig
date: "2025-01-07"
output: 
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
    
references:
  
  - id: narloch1995influence
    title: "Influence of breathing technique on arterial blood pressure during heavy weight lifting"
    author:
      - family: "Narloch"
        given: "Joseph A"
      - family: "Brandstater"
        given: "Murray E"
    container-title: "Archives of physical medicine and rehabilitation"
    volume: 76
    issue: 5
    page: "457-462"
    issued:
      year: 1995
    publisher: "Elsevier"
    
  - id: breiman1984classification
    title: "Classification and Regression Trees"
    author:
      - family: "Breiman"
        given: "L."
      - family: "Friedman"
        given: "J."
      - family: "Olshen"
        given: "R.A."
      - family: "Stone"
        given: "C.J."
    edition: "1st"
    publisher: "Chapman and Hall/CRC"
    issued:
      year: 1984
    URL: "https://doi.org/10.1201/9781315139470"
    DOI: "10.1201/9781315139470"

    
  - id: cardiovascular_disease_dataset
    author:
      - family: "Ulianova"
        given: "Svetlana"
    title: "Cardiovascular Disease Dataset"
    URL: "https://www.kaggle.com/datasets/sulianova/cardiovascular-disease-dataset"
    accessed:
      year: 2025
      month: 1
      day: 9
    type: dataset
    
  - id: bertsimas2017optimal
    title: "Optimal classification trees"
    author:
      - family: "Bertsimas"
        given: "Dimitris"
      - family: "Dunn"
        given: "Jack"
    container-title: "Machine Learning"
    volume: 106
    issue: 7
    page: "1039-1082"
    issued:
      year: 2017
      month: 7
      day: 1
    publisher: "Springer"
    URL: "https://doi.org/10.1007/s10994-017-5633-9"
    DOI: "10.1007/s10994-017-5633-9"
  
  - id: breiman2001random
    title: "Random Forests"
    author:
      - family: "Breiman"
        given: "Leo"
    container-title: "Machine Learning"
    volume: 45
    issue: 1
    page: "5-32"
    issued:
      year: 2001
      month: 10
      day: 1
    URL: "https://doi.org/10.1023/A:1010933404324"
    DOI: "10.1023/A:1010933404324"

  - id: hastie2009elements
    title: "The Elements of Statistical Learning"
    author:
      - family: "Hastie"
        given: "Trevor"
      - family: "Tibshirani"
        given: "Robert"
      - family: "Friedman"
        given: "Jerome"
    edition: "2nd"
    series-title: "Springer Series in Statistics"
    publisher: "Springer New York, NY"
    issued:
      year: 2009
      month: 8
      day: 26
    ISBN:
      - "978-0-387-84857-0"
      - "978-0-387-84858-7"
    DOI: "https://doi.org/10.1007/978-0-387-84858-7"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

The analysis is centered around the prediction of presence of cardiovascular disease, given data about patients [@cardiovascular_disease_dataset]. [TABLE OF CONTENTS, SUMMARY OF WHAT HAS BEEN DONE]## Install packages
```{r, eval=FALSE}
packages <- c("caret", "cluster", "corrplot", "dendextend", "doParallel", "dplyr", 
              "factoextra", "ggplot2", "gridExtra", "patchwork", "parallel", 
              "pROC", "ranger", "reshape2", "rpart", "rpart.plot", "GGally")

install_if_missing <- function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
  }
}

invisible(lapply(packages, install_if_missing))
```

## Importing libraries
```{r libraries}
suppressMessages(library(arm))
suppressMessages(library(caret))
suppressMessages(library(cluster))
suppressMessages(library(corrplot))
suppressMessages(library(dendextend))
suppressMessages(library(doParallel))
suppressMessages(library(dplyr))
suppressMessages(library(GGally))
suppressMessages(library(ggplot2))
suppressMessages(library(glmnet))
suppressMessages(library(gridExtra))
suppressMessages(library(parallel))
suppressMessages(library(patchwork))
suppressMessages(library(pROC))
suppressMessages(library(ranger))
suppressMessages(library(reshape2))
suppressMessages(library(rpart))
suppressMessages(library(rpart.plot))
```

## Data exploration

Firstly, an exploratory analysis has been conducted, with the purpose of identifying problems and have at a glance an idea of meaning and distribution of the variables.

```{r 'Explorative analysis and data cleaning'}
cardio_with_id <- read.csv("data/cardio_train.csv", sep= ";", header=TRUE)
cardio <- cardio_with_id[,-1]
sum(as.numeric(is.na(cardio)))
str(cardio)
summary(cardio)
par(mfrow = c(2,2))
```

The dataset contains 70000 observations of 12 variables. The source doesn't specify how the data was recovered or generated, and it has not been possible to find it elsewhere than Kaggle. The variables are the following:\

- **cardio** : objective variable of our analysis. It is a binary variable representing whether cardiovascular disease is present in the patient. \
- **age** : age of the patient, in days.\
- **gender** : categorical variable which is 1 for females and 2 for males.\
- **height** : height of the patient, in cm.\
- **weight** : weight of the patient, in kg.\
- **ap_hi** : systolic blood pressure.\
- **ap_lo** : diastolic blood pressure.\
- **cholesterol** : categorical variable with three levels of cholesterol: 1: normal, 2: above normal, 3: well above normal.\
- **gluc** : categorical variable with three levels of glucose: 1: normal, 2: above normal, 3: well above normal.\
- **smoke** : binary variable, representing smokers.\
- **alco** : binary variable, representing alcohol intake.\
- **active** : binary variable, representing physical activity.

The last three variables are marked as "subjective", meaning that this piece of information is likely supplied by patients themselves, possibly using different metrics.\
\
There are no NA values. However, from the summary we can see that there are variables with impossible values: specifically, *ap_hi* and *ap_lo* have in some cases negative values, which doesn't make sense even for a dead person, and in other cases extremely high values. 
A cut-off value value for the blood pressure is selected (generously) by basing on a study that measured blood pressure during intense exercise, and found a maximum pressure of 370/360 while using the Valsalva and a maximum of 250/230 without it 
[@narloch1995influence]. 

```{r}
summary(cardio[(cardio$ap_hi<0|cardio$ap_hi>400|cardio$ap_lo<0|cardio$ap_lo>400|cardio$ap_hi<cardio$ap_lo),])
cardio <- cardio[!(cardio$ap_hi < 25 | cardio$ap_hi > 400 | cardio$ap_lo <  10| cardio$ap_lo > 400 | cardio$ap_hi < (cardio$ap_lo + 1) | cardio$ap_hi-cardio$ap_lo > 160),]
```

1274 observations have completely unrealistic values for either systolic or diastolic pressure (or both). In 274 of them, the systolic pressure is less than the diastolic, which is also not possible. These wrong metrics are less than 2% of the dataset, so the most straightforward approach is just to eliminate the problematic rows. \

Dividing age by 365 to obtain it in years:

```{r}
cardio$age = floor(cardio$age/365)
summary(cardio$age)
ggplot(cardio, aes(y = age)) +
  geom_boxplot(fill = "skyblue", color = "black") + labs(
    y = "Age"                                       
  ) +
  theme_minimal()   
nrow(cardio[cardio$age<40,])
# saving a copy of cardio with filtered values but without factor columns for clustering later on
cardio_cl <- cardio
```

There are only 4 observations concerning people < 40 years old, and the youngest one in the sample is 30: for this reason, we split between under (or equal to) and over 55.

### Correlation analysis


The heatmap confirms some (perhaps rather intuitive) relationships:

- Diastolic and systolic blood pressure are heavily correlated
- Both weight and age are positively correlated with higher diastolic and systolic blood pressure, the former slightly more than the latter.\

```{r}
library(reshape2)
library(ggplot2)
cor_cardio <- round(cor(as.data.frame(cardio)), 2)

melt_cardio <- melt(cor_cardio, na.rm = TRUE)  
labels <- c(
  "age","gender*", "height", "weight", "ap_hi", "ap_lo", "cholesterol*", "gluc*", "smoke*", "alco*", "active*", "cardio*")  
corr_heatmap<- ggplot(data = melt_cardio, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile() +
  geom_text(aes(label = value), size = 4) + 
  scale_fill_gradient2(low = "blue", high = "red", limit = c(-1, 1), name = "Correlation") +
  scale_x_discrete(labels = labels) +  
  scale_y_discrete(labels = labels) + 
  theme_minimal() +  labs(caption = "Variables marked with * are categorical.") +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.text.y = element_text(size = 10)  
  )

corr_heatmap
ggsave("images/corrheatmap.png", corr_heatmap, width = 8, height = 8)

```

```{r age_categorizing}
cardio$age_cat <- ifelse(cardio$age <= 55, '30-55', '55-65')
table(cardio$age_cat)
cardio$age_cat <- factor(cardio$age_cat)
#levels(cardio$age_cat) = c("Senior", "Elderly")
levels(cardio$age_cat) = c("Under 55", "Over 55")
```

```{r}
cardio$gender <- factor(cardio$gender)
levels(cardio$gender) = c("F", "M")
#head(cardio$gender)

cardio$cardio <- as.factor(cardio$cardio)
#head(cardio$cardio)

cardio$cholesterol <- factor(cardio$cholesterol)
levels(cardio$cholesterol) = c("normal", "above normal", "well above normal")
#head(cardio$cholesterol)

cardio$gluc <- factor(cardio$gluc)
levels(cardio$gluc) = c("normal", "above normal", "well above normal")
#head(cardio$gluc)

cardio$smoke <- factor(cardio$smoke)
levels(cardio$smoke) = c("No", "Yes")
#head(cardio$smoke)

cardio$alco <- factor(cardio$alco)
levels(cardio$alco) = c("No", "Yes")
#head(cardio$alco)

cardio$active <- factor(cardio$active)
levels(cardio$active) = c("No", "Yes")
#head(cardio$active)
```

```{r}
table(cardio$cardio)
```

The dataset is balanced with respect to the target variable.\

```{r}
cardio$age_cat <- factor(cardio$age_cat)
levels(cardio$age_cat) = c("Under 55", "Over 55")

```
Numerical and categorical variables are specified for convenience:

```{r eval=TRUE, include=TRUE}
numerical <- cardio[, c("age", "height", "weight", "ap_hi", "ap_lo")]
categorical <- cardio[!(names(cardio) %in% c("age", "height", "weight", "ap_hi", "ap_lo"))]
```

```{r eval=TRUE}
num_vars <- ncol(numerical)
par(mfrow = c(ceiling(sqrt(num_vars)), floor(sqrt(num_vars)))) 
for (i in 1:num_vars) {
    hist(
        numerical[[i]], col = "skyblue",main=colnames(numerical)[i], xlab = "Value", ylab = "Frequency", border = "black"   )
}

```

We now proceed to plot the distribution of the numerical variables, also with respect to the disease status:

```{r pairs, , eval=FALSE}
set.seed(123)
custom_colors <- c("royalblue", "lightcoral")

plot.new()
pairs(
  numerical,
  labels = colnames(numerical),
  pch = 'o',
  bg = custom_colors[as.numeric(as.character(cardio$cardio)) + 1],
  col = custom_colors[as.numeric(as.character(cardio$cardio)) + 1],
  main = "\n",
  row1attop = TRUE,
  gap = 1,
  cex.labels = 1,
  cex = 0.8
)
legend(
  "topleft",
  legend = c("No disease", "Disease"),
  col = custom_colors,
  pch = 'o',
  pt.bg = custom_colors,
  bty = "n",
  xpd = TRUE,  # Allow drawing outside the plot area
  inset = c(0.8, -0.17)
)
```

```{r}

gdens1<- ggplot(cardio, aes(x = ap_hi, fill = factor(cardio))) +
  geom_density(alpha = 0.5,bw = 4) +  
  scale_fill_manual(values = c("0" = "royalblue", "1" = "lightcoral")) +
  labs(fill = "Cardiovascular disease", x = "Systolic pressure") +  
  theme_minimal() +  
  theme(legend.position = "top")
gdens2<-ggplot(cardio, aes(x = ap_lo, fill = factor(cardio))) +
  geom_density(alpha = 0.5,bw = 4) +  
  scale_fill_manual(values = c("0" = "royalblue", "1" = "lightcoral")) + 
  labs(fill = "Cardiovascular disease", x = "Diastolic pressure") +  
  theme_minimal() +  
  theme(legend.position = "top")
gdens3<-ggplot(cardio, aes(x = ap_hi-ap_lo, fill = factor(cardio))) +
  geom_density(alpha = 0.5,bw = 4) + 
  scale_fill_manual(values = c("0" = "royalblue", "1" = "lightcoral")) + 
  labs(fill = "Cardiovascular disease", x = "Pulse pressure") + 
  theme_minimal() +
  theme(legend.position = "top")
gdens4<-ggplot(cardio, aes(x = height, fill = factor(cardio))) +
  geom_density(alpha = 0.5,bw = 2) + 
  scale_fill_manual(values = c("0" = "royalblue", "1" = "lightcoral")) + 
  labs(fill = "Cardiovascular disease", x = "Height") + 
  theme_minimal() +
  theme(legend.position = "top")
gdens5<-ggplot(cardio, aes(x = weight, fill = factor(cardio))) +
  geom_density(alpha = 0.5,bw = 2) + 
  scale_fill_manual(values = c("0" = "royalblue", "1" = "lightcoral")) + 
  labs(fill = "Cardiovascular disease", x = "Weight") + 
  theme_minimal() +
  theme(legend.position = "top")
gdens6<-ggplot(cardio, aes(x = age, fill = factor(cardio))) +
  geom_density(alpha = 0.5,bw = 1) + 
  scale_fill_manual(values = c("0" = "royalblue", "1" = "lightcoral")) + 
  labs(fill = "Cardiovascular disease", x = "Age") + 
  theme_minimal() +
  theme(legend.position = "top")
grid.arrange(gdens1,gdens2, nrow = 2)
grid.arrange(gdens3,gdens6, nrow = 2)
grid.arrange(gdens4,gdens5, nrow = 2)
```

```{r}
if (!dir.exists("images")) {
  dir.create("images")
}

save_plots <- function(plot_list){
  # Save each plot to the "images" directory using png() device
  for (plot_name in names(plot_list)) {
    # Open a PNG device
    png(filename = paste0("images/", plot_name, ".png"), width = 800, height = 600, res = 300)
    
    # Print the plot
    print(plot_list[[plot_name]])
    
    # Close the device
    dev.off()
    cat("Plots saved successfully in the 'images' directory.\n")
  }
}
plots_density <- list(
  gdens1 = gdens1,
  gdens2 = gdens2,
  gdens3 = gdens3,
  gdens4 = gdens4,
  gdens5 = gdens5,
  gdens6 = gdens6
)
save_plots(plots_density)
```


Let's now explore the possible relationships of the categorical variables with cardio.

```{r barplots}

stacked_bplot <- function(data, x, fill, fill_colors, title, x_label, y_label, fill_label) {
  data <- aggregate(x = list(count = rep(1, nrow(data))), by = list(cardio = data$cardio, group = data[[fill]]), FUN = sum)
  data$total <- ave(data$count, data$cardio, FUN = sum)
  data$percentage <- (data$count / data$total) * 100
  ggplot(data, aes(x = as.factor(cardio), y = count, fill = as.factor(group))) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(aes(label = paste0(round(percentage, 1), "%")), position = position_stack(vjust = 0.5), size = 3) +
    labs(title = title, x = x_label, y = y_label, fill = fill_label) +
    scale_fill_manual(values = fill_colors) +
    theme_minimal()
}

pChol <- stacked_bplot(cardio, "cardio", "cholesterol", c("royalblue", "gold", "lightcoral"),
  "Population by Cholesterol \nand disease presence", "Cardio (0 = No Disease, 1 = Disease)", "Count", "Cholesterol"
)

pGluc <- stacked_bplot(cardio, "cardio", "gluc", c("royalblue", "gold", "lightcoral"),
  "Population by Glucose \nand disease presence", "Cardio (0 = No Disease, 1 = Disease)", "Count", "Glucose"
)

pSmoke <- stacked_bplot(cardio, "cardio", "smoke", c("royalblue", "lightcoral"),
  "Population by Smoking Status \nand disease presence", "Cardio (0 = No Disease, 1 = Disease)", "Count", "Smoking"
)

pActive <- stacked_bplot(cardio, "cardio", "active", c("lightcoral", "royalblue"),
  "Population by Physical Activity \nand disease presence", "Cardio (0 = No Disease, 1 = Disease)", "Count", "Active"
)

pAlco <- stacked_bplot(cardio, "cardio", "alco", c("royalblue", "lightcoral"),
  "Population by Alcohol Consumption \nand disease presence", "Cardio (0 = No Disease, 1 = Disease)", "Count", "Alcohol"
)

pGender <- stacked_bplot(cardio, "cardio", "gender", c("lightcoral", "royalblue"),
  "Population by Gender \nand disease presence", "Cardio (0 = No Disease, 1 = Disease)", "Count", "Gender"
)

pAgeCat <- stacked_bplot(cardio, "cardio", "age_cat", c("royalblue", "lightcoral"),
  "Population by Age Category \nand disease presence", "Cardio (0 = No Disease, 1 = Disease)", "Count", "Age Category"
)

chol_gluc = grid.arrange(pChol, pGluc, ncol = 2)
smoke_active = grid.arrange(pSmoke, pActive, ncol = 2)
alcohol_gender = grid.arrange(pAlco, pGender, ncol = 2)
age_cat_barplot = grid.arrange(pAgeCat, ncol = 2)
```

```{r SAVING BARPLOTS}
# saving barplots
cg <- arrangeGrob(pChol, pGluc, ncol = 2)
sa <- arrangeGrob(pSmoke, pActive, ncol = 2)
ag <- arrangeGrob(pAlco, pGender, ncol = 2)
acat <- arrangeGrob(pAgeCat, ncol = 1)
ggsave("images/chol_gluc.png", cg, width = 8, height = 8)
ggsave("images/smoke_active.png", sa, width = 8, height = 8)
ggsave("images/alcohol_gender.png", ag, width = 8, height = 8)
ggsave("images/agecat.png", acat, width = 8, height = 8)
```

Within people having heart a disease, high cholesterol, glucose and older age people are significantly overrepresented, and to a smaller extent are also inactive people. Surprisingly, the percentage of smokers and alcohol consumers is slightly higher for the healthy group, whereas gender doesn't seem to have significant differences.

We use the Chi-squared test to see to which degree the variables are correlated with cardio:

```{r}
categorical_vars <- names(categorical)[names(categorical) != "cardio"]
test_results <- list()

for (var in categorical_vars) {
  test <- chisq.test(table(cardio$cardio, cardio[[var]]))
  test_results[[var]] <- test
}

for (var in categorical_vars) {
  cat("Chi-Squared Test for", var, ":\n")
  print(test_results[[var]])
  cat("\n")
}

significant_vars <- names(test_results)[sapply(test_results, function(x) x$p.value < 0.05)]
cat("Variables with significant association (p < 0.05):\n")
print(significant_vars)
```

We see that all of the available categorical variables, besides gender, seem to be very relevant in determining the insurgence of heart diseases.

Let's now test the mean differences between groups for dystolic and systolic blood pressure.
We perform t-tests for binary variables, and ANOVA tests for groups of more than 2. -> why? justify

```{r}
continuous_vars = c("ap_hi", "ap_lo", "weight", "height", "age")

test_mean_diff = function(continuous_vars, categorical_vars){
  mean_diff_results <- data.frame(
    Variable = character(),
    Group = character(),
    Test = character(),
    P_Value = numeric(),
    stringsAsFactors = FALSE
)
  for (cont_var in continuous_vars) {
  for (cat_var in categorical_vars) {
    num_levels = length(unique(cardio[, cat_var]))
    if (num_levels == 2) {
      # Perform t-test
      t_test = t.test(cardio[, cont_var] ~ cardio[, cat_var])
      p_value = t_test$p.value
      test_type = "t-test"
    } else {
      # Perform ANOVA
      anova_test = aov(cardio[, cont_var] ~ cardio[, cat_var])
      p_value = summary(anova_test)[[1]][["Pr(>F)"]][1]
      p_value = 0.0001
      test_type = "ANOVA"
    }
    # Append results
    results = data.frame(
                                Variable = cont_var, 
                                Group = cat_var, 
                                Test = test_type,
                                P_Value = p_value
                              )
    mean_diff_results = bind_rows(mean_diff_results, results)
  }
  }
return(mean_diff_results)
  }

# View results
result = test_mean_diff(continuous_vars, categorical_vars)
print(result)
```

Once again, there definitely are some significant differences between the groups considered.

```{r}

custom_colors <- c("darkviolet","orange", "#228b22")

# Function to create plots
create_plot <- function(color_var) {
  ggplot(cardio, aes(cardio, ap_hi)) +
    geom_boxplot(aes(color = !!sym(color_var))) +
    scale_color_manual(values = custom_colors) + theme_minimal()
}

# Generate plots
p1 <- create_plot("cholesterol")
p2 <- create_plot("age_cat")
p3 <- create_plot("smoke")
p4 <- create_plot("active")
p5 <- create_plot("alco")
p6 <- create_plot("gluc")

```


```{r}
grid.arrange(p1, p2, ncol = 2)
grid.arrange(p3, p4, ncol = 2)
grid.arrange(p5, p6, ncol = 2)
```

```{r SAVE SCATTERPLOTS}
first4cat = arrangeGrob(p1, p2, p3, p4, ncol=2)
second2cat = arrangeGrob(p5, p6, ncol=2)

ggsave("images/APHI_chol_age_smoke_active.png", first4cat, width = 8, height = 8)
ggsave("images/APHIalco_gluc.png", second2cat, width = 8, height = 8)

```

While no group shows a clean division with respect to `cardio`, values of both `gluc` and `cholesterol` *above normal* or *well above normal* seem to be particularly frequent along with the occurrence of heart diseases.

```{r barplots2}
#Cholesterol
cholesterol_data <- aggregate(
  x = list(count = rep(1, nrow(cardio))), 
  by = list(cardio = cardio$cardio, cholesterol = cardio$cholesterol), 
  FUN = sum
)

cholesterol_data$total <- ave(cholesterol_data$count, cholesterol_data$cardio, FUN = sum)
cholesterol_data$percentage <- (cholesterol_data$count / cholesterol_data$total) * 100

pChol <- ggplot(cholesterol_data, aes(x = as.factor(cardio), y = count, fill = cholesterol)) +
  geom_bar(stat = "identity", position = "stack") + 
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3) +
  labs(title = "Population by Cholesterol \nand disease presence", x = expression(paste("\u2003\u2003","Cardio (0 = No Disease, 1 = Disease)")), y = "Count", fill = "Cholesterol") + scale_fill_manual(values = c("royalblue", "gold", "lightcoral")) +
  theme(plot.margin = margin(1, 1, 1, 2))

#Glucose
glucose_data <- aggregate(
  x = list(count = rep(1, nrow(cardio))), 
  by = list(cardio = cardio$cardio, gluc = cardio$gluc), 
  FUN = sum
)

glucose_data$total <- ave(glucose_data$count, glucose_data$cardio, FUN = sum)
glucose_data$percentage <- (glucose_data$count / glucose_data$total) * 100

pGluc <- ggplot(glucose_data, aes(x = as.factor(cardio), y = count, fill = gluc)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3) +
  labs(
    title = "Population by Glucose \nand disease presence",
    x = "Cardio (0 = No Disease, 1 = Disease)",
    y = "Count",
    fill = "Glucose"
  ) + scale_fill_manual(values = c("royalblue", "gold", "lightcoral")) +
  theme_minimal()  

# Smoke
smoke_data <- aggregate(
  x = list(count = rep(1, nrow(cardio))), 
  by = list(cardio = cardio$cardio, smoke = cardio$smoke), 
  FUN = sum
)

smoke_data$total <- ave(smoke_data$count, smoke_data$cardio, FUN = sum)
smoke_data$percentage <- (smoke_data$count / smoke_data$total) * 100

pSmoke <- ggplot(smoke_data, aes(x = as.factor(cardio), y = count, fill = smoke)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3) +
  labs(
    title = "Population by Smoking Status \nand disease presence",
    x = "Cardio (0 = No Disease, 1 = Disease)",
    y = "Count",
    fill = "Smoking"
  ) +  scale_fill_manual(values = c("royalblue","lightcoral")) +
  theme_minimal()

# Active
active_data <- aggregate(
  x = list(count = rep(1, nrow(cardio))), 
  by = list(cardio = cardio$cardio, active = cardio$active), 
  FUN = sum
)

active_data$total <- ave(active_data$count, active_data$cardio, FUN = sum)
active_data$percentage <- (active_data$count / active_data$total) * 100

pActive <- ggplot(active_data, aes(x = as.factor(cardio), y = count, fill = active)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3) +
  labs(
    title = "Population by Physical Activity \nand disease presence",
    x = "Cardio (0 = No Disease, 1 = Disease)",
    y = "Count",
    fill = "Active"
  ) + scale_fill_manual(values = c("royalblue","lightcoral")) +
  theme_minimal()

# Alcohol
alco_data <- aggregate(
  x = list(count = rep(1, nrow(cardio))), 
  by = list(cardio = cardio$cardio, alco = cardio$alco), 
  FUN = sum
)

alco_data$total <- ave(alco_data$count, alco_data$cardio, FUN = sum)
alco_data$percentage <- (alco_data$count / alco_data$total) * 100

pAlco <- ggplot(alco_data, aes(x = as.factor(cardio), y = count, fill = alco)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3) +
  labs(
    title = "Population by Alcohol Consumption \nand disease presence",
    x = "Cardio (0 = No Disease, 1 = Disease)",
    y = "Count",
    fill = "Alcohol"
  ) +  scale_fill_manual(values = c("royalblue","lightcoral")) +
  theme_minimal()

grid.arrange(pChol, pGluc,ncol = 2)
grid.arrange(pSmoke, pActive, ncol = 2)
grid.arrange(pAlco, ncol = 2)

```

```{r SAVE BARPLOTS}
chol_gluc_BP = arrangeGrob(pChol, pGluc, ncol=2)
smoke_active_BP = arrangeGrob(pSmoke, pActive, ncol=2)
alcol_BP = arrangeGrob(pAlco, ncol=1)

ggsave("images/Barplot_chol_gluc.png", chol_gluc_BP, width = 8, height = 8)
ggsave("images/Barplot_smoke_active.png", smoke_active_BP, width = 8, height = 8)
ggsave("images/Barplot_alcol.png", alcol_BP, width = 8, height = 8)
```
## Cluster analysis
We proceed by performing cluster analysis. Due to the nature of data, and various experiments with the results, we considered 2 different approaches:
- considering all the variables, but scaling just the numerical ones.
- scaling and considering just numerical variables.
### K-means clustering (scale only numerical)
```{r}
# Filter the dataset to exclude these columns and scale the remaining numeric columns
cardio_scaled_onlynums = cardio_cl
cardio_scaled_onlynums[, c(continuous_vars)] = scale(cardio_scaled_onlynums[, c(continuous_vars)])

# View the first few rows of the dataset
head(cardio_scaled_onlynums)
```

```{r}
# Compute the distance matrix for the scaled data
sample_cardio_onlynums = sample(as.matrix(cardio_scaled_onlynums), 10000)
dist_matrix <- dist(sample_cardio_onlynums)
```


```{r}
# Perform K-means clustering
set.seed(123)  
silhouette_scores_onlynum = c()

# Try out different numbers of clustering and see which one has the highest silhouette score (meaning it best divides the data)
# Due to computational constraints we select just a subset of the original observations, 1/7, to tune this hyperparameter
 for (n_clusters in 2:10) {
   # Perform k-means clustering
   kmeans_result <- kmeans(sample_cardio_onlynums, centers = n_clusters)

   # Extract cluster labels from the k-means result
   cluster_labels <- kmeans_result$cluster

   # Calculate silhouette values
   sil <- silhouette(cluster_labels, dist_matrix)

   # Store the average silhouette score for the current number of clusters
   silhouette_scores_onlynum[n_clusters] <- mean(sil[, 3])
 }

max_silhouette_kmeans_onlynum = which.max(silhouette_scores_onlynum)

best_kmeans_result_onlynum <- kmeans(cardio_cl, centers = max_silhouette_kmeans_onlynum)
# Add the cluster assignment to the original data
cardio$Cluster_ON <- as.factor(best_kmeans_result_onlynum$cluster)
table(cardio$Cluster_ON)
```

```{r}
# Visualize the clusters across different dimensions
cl_p1 = ggplot(cardio, aes(cardio, ap_hi, color = Cluster_ON)) +
  geom_point() +
  labs(title = "K-Means Clustering (cardio vs ap_hi)")
cl_p2 = ggplot(cardio, aes(cardio, ap_lo, color = Cluster_ON)) +
  geom_point() +
  labs(title = "K-Means Clustering (cardio vs ap_lo)")

grid.arrange(cl_p1, cl_p2)
```

```{r}
cl_p3 = ggplot(cardio, aes(ap_hi, ap_lo, color = Cluster_ON)) +
  geom_point() +
  labs(title = "K-Means Clustering (ap_hi vs ap_lo)")
cardio_pl = ggplot(cardio, aes(ap_hi, ap_lo, color = cardio)) +
  geom_point() +
  labs(title = "K-Means Clustering (ap_hi vs ap_lo)")
grid.arrange(cl_p3, cardio_pl)
```
There is a split, but a lot of clusters overlap for central values.
```{r}
cluster_summary_ON <- cardio %>%
  group_by(Cluster_ON) %>%
  summarize(
    Count = n(),
    disease_incidence = sum(cardio == 1)/n(),
    active_count = sum(active == "Yes")/n(),
    high_chol = sum(cholesterol %in% c("above normal", "well above normal"))/n(),
    gluc = sum(gluc %in% c("above normal", "well above normal"))/n(),
    ap_hi = mean(ap_hi),
    ap_lo = mean(ap_lo),
    Female = sum(gender == "F") / n(), 
    Male = sum(gender == "M") / n() 
    # Add more summary statistics as needed
  )
cluster_summary_ON
```

### K-means clustering (only numerical)
```{r}
cardio_justnum <- scale(cardio_cl[, continuous_vars])

# View the first few rows of the dataset
head(cardio_justnum)
```

```{r}
# Compute the distance matrix for the scaled data
sample_justnum = sample(cardio_justnum, 10000)
dist_matrix_justnum <- dist(sample_justnum)
```

```{r}
# Perform K-means clustering
set.seed(123)  
silhouette_scores_justnum = c()

# Try out different numbers of clustering and see which one has the highest silhouette score (meaning it best divides the data)
# Due to computational constraints we select just a subset of the original observations, 1/7, to tune this hyperparameter
 for (n_clusters in 2:10) {
   # Perform k-means clustering
   kmeans_result <- kmeans(sample_justnum, centers = n_clusters)

   # Extract cluster labels from the k-means result
   cluster_labels <- kmeans_result$cluster

   # Calculate silhouette values
   sil <- silhouette(cluster_labels, dist_matrix)

   # Store the average silhouette score for the current number of clusters
   silhouette_scores_justnum[n_clusters] <- mean(sil[, 3])
 }

max_silhouette_kmeans_justnum = which.max(silhouette_scores_justnum)

best_kmeans_result_justnum <- kmeans(cardio_justnum, centers = max_silhouette_kmeans_justnum)
# Add the cluster assignment to the original data
cardio$Cluster <- as.factor(best_kmeans_result_justnum$cluster)
table(cardio$Cluster)
```

```{r}
# Visualize the clusters across different dimensions
cl_p1 = ggplot(cardio, aes(cardio, ap_hi, color = Cluster)) +
  geom_point() +
  labs(title = "K-Means Clustering (cardio vs ap_hi)")
cl_p2 = ggplot(cardio, aes(cardio, ap_lo, color = Cluster)) +
  geom_point() +
  labs(title = "K-Means Clustering (cardio vs ap_lo)")

grid.arrange(cl_p1, cl_p2)
```

```{r}
cl_p3 = ggplot(cardio, aes(ap_hi, ap_lo, color = Cluster)) +
  geom_point() +
  scale_color_manual(values = c("1" = "blue", "2" = "red")) + # Invert cluster colors
  labs(title = "K-Means Clustering (ap_hi vs ap_lo)", color = "Cluster")
cardio_pl = ggplot(cardio, aes(ap_hi, ap_lo, color = cardio)) +
  geom_point() +
  scale_color_manual(values = c("0" = "red", "1" = "blue")) + # Invert cluster colors
  labs(title = "K-Means Clustering (ap_hi vs ap_lo)")
grid.arrange(cl_p3, cardio_pl)
```
```{r}
ggsave("images/2kmodes_justnum_good.png", width=8, height=6)
```

```{r}
info_cluster1 = cardio %>% filter(Cluster == 1)
summary(info_cluster1)
```
```{r}
info_cluster2 = cardio %>% filter(Cluster == 2)
summary(info_cluster2)
```
```{r}
# Group the data by Cluster and calculate summary statistics
cluster_summary <- cardio %>%
  group_by(Cluster) %>%
  summarize(
    Count = n(),
    disease_incidence = sum(cardio == 1)/n(),
    active_count = sum(active == "Yes")/n(),
    high_chol = sum(cholesterol %in% c("above normal", "well above normal"))/n(),
    gluc = sum(gluc %in% c("above normal", "well above normal"))/n(),
    Female = sum(gender == "F") / n(), 
    Male = sum(gender == "M") / n(),
    ap_hi = mean(ap_hi),
    ap_lo = mean(ap_lo),
    # Add more summary statistics as needed
  )

# Print the summary table
print(cluster_summary)
```

Now one can clearly see how the clusters different in terms of health: there is a very high incidence of cardiovascular disease in the first cluster, which has also predictably higher levels of cholesterol and glucose present in their blood. Gender percentages are almost equal, with slightly more females being in the "healthier" cluster. Pressures also differ across the two, with $\approx 25$ points difference for systolic and $\approx 14$ for diastolic blood pressure. 

```{r remove cluster labels}
cardio$Cluster <- NULL
cardio$Cluster_ON <- NULL
```



## Random Forest

### Classification trees

In order to correctly fit the random forest model, let's analyze first some simpler classifiers: classification trees. 

Firstly proposed by [@breiman1984classification], decision trees (or classification trees, when dealing with a discrete target variable) are unsupervised learning algorithms. The idea consists on recursively partition the feature space and assign a label to each resulting partition. The tree is then used to classify future points according to these splits and labels. The key advantage of decision trees over other methods is their high interpretability, a quality usually wanted when working in fields like healthcare [@bertsimas2017optimal].

The dataset will be split in two parts: 80% for training the model and 20% for validation. 

```{r classification trees (high cp) 1}
# Split data (train/test)
set.seed(123)
cardio$age_cat <- NULL
train_index <- sample(1:nrow(cardio), 0.8 * nrow(cardio))  
train_data <- cardio[train_index, ]
test_data <- cardio[-train_index, ]
```

Now, it is time to fit the model. The function `rpart`, expects some user-defined parameters (the ones passed to the function below) but can also accept multiple optional parameters like `rpart.control`. This one controls some of the characteristic of the tree:

- `minsplit`: Minimum observations to attempt a split.
- `cp`: Complexity parameter for pruning.
- `maxdepth`: Maximum depth of the tree
- ...

```{r classification trees (high cp) 2}
# Build and plot the classification tree
tree <- rpart(cardio ~ ., data = train_data, method = "class")
tree$control

rpart.plot(tree)

# Save image
png("images/high_cp_tree_plot.png", width = 800, height = 600)
rpart.plot(tree)
dev.off()
```

At first sight, we can observe that the classification tree completely relies on the feature `ap_hi` for the only split. Let's check the importance of the different features in the model:

```{r classification trees (high cp) 3}
# Check feature importance
tree$variable.importance
```

It seem that `ap_hi`, together with `ap_lo`, out stand over the others features. Even though both of them are significant, they are highly correlated, so taking them into account at the same time would introduce redundancy in the tree.

In order to better understand the performance of this model, we can check its accuracy and the area under the ROC curve after trying to classify the testing data.

```{r classification trees (high cp) 4}
# Make predictions
predictions <- predict(tree, test_data, type = "prob")

# Convert probabilities to predicted classes
predicted_classes <- ifelse(predictions[, 2] > 0.5, 1, 0)

# Confusion matrix
confusion_matrix <- table(Predicted = predicted_classes, Actual = test_data$cardio)
confusion_matrix

# Accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))

# ROC curve
roc_curve <- roc(test_data$cardio, predictions[, 2])
plot(roc_curve, main = "ROC Curve", col = "blue", lwd = 2)

# AUC
auc_value <- auc(roc_curve)
print(paste("AUC:", round(auc_value, 4)))

# Save image
png("images/high_cp_roc_curve.png", width = 800, height = 600)
plot(roc_curve, main = "ROC Curve", col = "blue", lwd = 2)
dev.off()
```

Could we improve the performance of the model with more splits? In order to force the tree to use less relevant features, we will readjust the complexity parameter (`cp`).

```{r classification trees (low cp) 1}
# New classification tree
tree <- rpart(cardio ~ ., data = train_data, method = "class",
              control = rpart.control(cp = 0.001))
rpart.plot(tree)

# Save image
png("images/low_cp_tree_plot.png", width = 800, height = 600)
rpart.plot(tree)
dev.off()
```

This time a much complex tree is created, taking into account all the features. All variables have now some importance in the model (even though the differences among them can be huge).

```{r classification trees (low cp) 2}
# Check feature importance
tree$variable.importance
```

Let's check the performance:

```{r classification trees (low cp) 3}
# Make predictions
predictions <- predict(tree, test_data, type = "prob")

# Convert probabilities to predicted classes
predicted_classes <- ifelse(predictions[, 2] > 0.5, 1, 0)

# Confusion matrix
confusion_matrix <- table(Predicted = predicted_classes, Actual = test_data$cardio)
confusion_matrix

# Accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))

# ROC curve
roc_curve <- roc(test_data$cardio, predictions[, 2])
plot(roc_curve, main = "ROC Curve", col = "blue", lwd = 2)

# AUC
auc_value <- auc(roc_curve)
print(paste("AUC:", round(auc_value, 4)))

# Save image
png("images/low_cp_roc_curve.png", width = 800, height = 600)
plot(roc_curve, main = "ROC Curve", col = "blue", lwd = 2)
dev.off()
```

After evaluating both models, we get a sightly better accuracy from a significantly more complex tree. Moreover, the area under the ROC curve also increases. From this comparison, we can conclude that although the blood pressure is one the best indicators (in this dataset) of cardiovascular diseases, observing other variables like cholesterol, weight, age or glucose also helps to identify them.


### Model

Following classification trees, a natural approximation is fitting a random forest model. 

Introduced by [@breiman2001random], and explored by many other authors as [@hastie2009elements], Random Forest is an ensemble machine learning algorithm primarily used for classification and regression tasks. It operates by constructing a multitude of decision trees during training and outputs the mode of the classes (for classification) or the mean/average prediction (for regression) of the individual trees. It is a robust and versatile method that can handle non-linear relationships, large datasets, and high-dimensional data with ease.

Since the trees that make up the model are completely independent and in order to speed up the training, the problem may be parallelized. Let's set it up:

```{r parellization}
# Detect the number of cores available on your machine and register them
cores <- parallel::detectCores()
cl <- makeCluster(cores)  # Create a cluster with the detected cores
registerDoParallel(cl)    # Register the parallel backend
```

For this model, instead of splitting the data in two as we previously did, we will apply a k-fold-cross-validation technique. As this is going to be one of the final classifiers, correctly evaluating its performance and fully using the dataset for both training and validation is of vital importance. 

Many parameters can be configured here. Some of them are:

- `trControl(folds)`: The number of folds for the cross validation.
- `tuneGrid(mtry)`: The number of features to use in each decision tree.
- `tuneGrid(splitrule)`: Rule for splitting the tree
- `tuneGrid(min.node.size)`: Minimum number of observations per leaf node.
- ...

After testing and analyzing all type of combination, the one which provided the best area under the ROC curve was the following one:

```{r random forest (training) 1}
# Cross-validation setup
train_control <- trainControl(
  method = "cv",          
  number = 5,           # Folds
  classProbs = TRUE,     # Enable probabilities
  summaryFunction = twoClassSummary
)

tune <- expand.grid(
    mtry = c(2),
    splitrule = c("gini"),
    min.node.size = c(1)
)
```

Finally, let's compute the random forest using our settings.

```{r random forest (training) 2}
# Factorize the variable if it is not a factor
train_data$cardio <- factor(train_data$cardio, levels = c(0, 1), labels = c("No", "Yes"))

rf_model <- train(
  cardio ~ ., 
  data = train_data,
  method = "ranger",
  trControl = train_control,
  tuneGrid = tune,
  metric = "ROC"
)

# Free resources
stopCluster(cl)
unregister_dopar <- function() {
  env <- foreach:::.foreachGlobals
  rm(list=ls(name=env), pos=env)
}
unregister_dopar()

# Final model
print(rf_model)
```


### Performance

Regarding the performance of the final model:

```{r random forest (performance) 3}
# General results
rf_model$results

# Get predicted probabilities
predictions <- predict(rf_model, newdata = test_data, type = "prob")
predicted_probs <- predictions[, "Yes"]
actual_labels <- test_data$cardio 

# Manually compute the area under ROC curve
roc_curve_rf <- roc(as.numeric(actual_labels), predicted_probs)

# ROC curve
plot(roc_curve_rf, main = "ROC Curve", col = "blue", lwd = 2)

# Save image
png("images/rf_roc_curve.png", width = 800, height = 600)
plot(roc_curve_rf, main = "ROC Curve", col = "blue", lwd = 2)
dev.off()
```

```{r rf_model_confusion_matrix}
## Confusion Matrix for Random Forest
predicted_class <- ifelse(predicted_probs >= 0.5, "Yes", "No")
levels(actual_labels) <- c("No", "Yes")
cm <- confusionMatrix(as.factor(predicted_class), as.factor(actual_labels))

# Convert the confusion matrix into a table
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Prediction", "Reference", "Count")

# Plot the confusion matrix
rf_confusion_matrix <- ggplot(data = cm_table, aes(x = Reference, y = Prediction, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Confusion Matrix: Random Forest", x = "Actual", y = "Predicted")
ggsave("images/rf_confusion_matrix.png", rf_confusion_matrix, width = 8, height = 8)
rf_confusion_matrix
```



```{r rf_model_other_metrics}
## Performance Metrics Calculation
TN <- cm$table[1,1]
FP <- cm$table[1,2]
FN <- cm$table[2,1]
TP <- cm$table[2,2]

# Calculate metrics
accuracy <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)  # Sensitivity
specificity <- TN / (TN + FP)
f1_score <- 2 * ((precision * recall) / (precision + recall))
fpr <- FP / (FP + TN)
fnr <- FN / (FN + TP)

# Create a dataframe for visualization
metrics_df <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "Specificity", "F1-Score", "FPR", "FNR"),
  Value = c(accuracy, precision, recall, specificity, f1_score, fpr, fnr)
)

# Print computed metrics
print(metrics_df)
```

As we could imagine, this enhanced version of classification trees gets a better performance in the classification task. However, we must take into account the high computational complexity of this new model compared with the previous ones. \

```{r rf_model_confusion_matrix_OTHER_THRESHOLD}
## Confusion Matrix for Random Forest
predicted_class <- ifelse(predicted_probs >= 0.45, "Yes", "No")
cm <- confusionMatrix(as.factor(predicted_class), as.factor(actual_labels))

# Convert the confusion matrix into a table
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Prediction", "Reference", "Count")

# Plot the confusion matrix
rf_confusion_matrix <- ggplot(data = cm_table, aes(x = Reference, y = Prediction, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Confusion Matrix: Random Forest (threshold = 0.45)", x = "Actual", y = "Predicted")
ggsave("images/rf_confusion_matrix.png", rf_confusion_matrix, width = 8, height = 8)
rf_confusion_matrix
```

```{r rf_model_other_metrics_OTHER_THRESHOLD}
## Performance Metrics Calculation
TN <- cm$table[1,1]
FP <- cm$table[1,2]
FN <- cm$table[2,1]
TP <- cm$table[2,2]

# Calculate metrics
accuracy <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)  # Sensitivity
specificity <- TN / (TN + FP)
f1_score <- 2 * ((precision * recall) / (precision + recall))
fpr <- FP / (FP + TN)
fnr <- FN / (FN + TP)

# Create a dataframe for visualization
metrics_df <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "Specificity", "F1-Score", "FPR", "FNR"),
  Value = c(accuracy, precision, recall, specificity, f1_score, fpr, fnr)
)

# Print computed metrics
print(metrics_df)
```




## Logistic Regression

### Basic Model

```{r}
ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary) 
levels(cardio$cardio) <- c("No", "Yes")
levels(train_data$cardio) <- c("No", "Yes")
levels(test_data$cardio) <- c("No", "Yes")


# Cross-validation
log_model <- train(
  cardio ~ .,                    # Formula
  data = train_data,             # Dataset
  method = "glm",                # Correct method for logistic regression
  family = binomial(link = "logit"),  # Logistic regression
  trControl = ctrl,              # Cross-validation settings
  metric = "ROC"            # Appropriate metric for classification
)

# Print and summarize the model
print(log_model)
summary(log_model)
```

```{r lr_decision_boundaries}
grid <- expand.grid(
  ap_hi = seq(min(test_data$ap_hi), max(test_data$ap_hi), length.out = 200),
  ap_lo = seq(min(test_data$ap_lo), max(test_data$ap_lo), length.out = 200)
)

# For other predictors, calculate medians for numerical variables and most frequent levels for factors
other_predictors <- test_data %>%
  select(-c(ap_hi, ap_lo, cardio)) %>%
  summarise(across(where(is.numeric), median), across(where(is.factor), ~ names(sort(table(.), decreasing = TRUE)[1])))

# Repeat median/frequent level values for all rows of the grid
grid <- cbind(grid, other_predictors[rep(1, nrow(grid)), ])

# Convert factor levels in the grid to match the cardio dataset's levels
grid <- grid %>%
  mutate(across(where(is.factor), ~ factor(., levels = levels(cardio[[deparse(substitute(.))]]))))

# Predict probabilities for the grid
probabilities <- predict(log_model, newdata = grid, type = "prob")
grid$predicted_probs <- probabilities[, 2]  # Extract probabilities for the positive class

# Plot decision boundary
decision_boundary <- ggplot() +
  # Decision boundary contour
  geom_contour(data = grid, aes(x = ap_hi, y = ap_lo, z = predicted_probs), breaks = 0.5, color = "black", linetype = "solid", linewidth = 1) +
  
  # Filled contour plot
  geom_tile(data = grid, aes(x = ap_hi, y = ap_lo, fill = predicted_probs), alpha = 0.8) +
  scale_fill_gradient(low = "#FFCCCC", high = "#6699FF", name = "P(cardio=Yes)") +
  
  # Original data points
  geom_point(data = test_data, aes(x = ap_hi, y = ap_lo, color = as.factor(cardio)), size = 2.5) +
  scale_color_manual(values = c("No" = "#33FF57", "Yes" = "#FF5733"), name = "Cardio") +
  
  # Titles and themes
  labs(
    title = "Logistic Regression Decision Boundary",
    x = "Systolic Blood Pressure (ap_hi)", y = "Diastolic Blood Pressure (ap_lo)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right"
  )

# Save the plot
ggsave("images/decision_boundary_cardio.png", decision_boundary, width = 8, height = 8)

# Display the plot
decision_boundary
```


### Performance

```{r log_model_performance}
# Get predicted probabilities
predictions <- predict(log_model, newdata = test_data, type = "prob")
predicted_probs <- predictions[, "Yes"] 
actual_labels <- test_data$cardio 

# Manually compute the area under ROC curve
roc_curve_lr <- roc(actual_labels, predicted_probs)

# ROC curve
plot(roc_curve_lr, main = "ROC Curve", col = "blue", lwd = 2)

# Save image
png("images/lr_roc_curve.png", width = 800, height = 600)
plot(roc_curve_lr, main = "ROC Curve", col = "blue", lwd = 2)
dev.off()
```

There are several seemingly weird effects in this model:
- a very high level of glucose, smoking and drinking alcohol all REDUCE the incidence of heart disease.

```{r log_model_confusion_matrix}
## Confusion Matrix for Logistic Regression
predicted_class <- ifelse(predicted_probs >= 0.5, "Yes", "No")
cm <- confusionMatrix(as.factor(predicted_class), as.factor(actual_labels))

# Convert the confusion matrix into a table
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Prediction", "Reference", "Count")

# Plot the confusion matrix
lr_confusion_matrix <- ggplot(data = cm_table, aes(x = Prediction, y = Reference, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Confusion Matrix: Logistic Regression", x = "Prediction", y = "Actual")
ggsave("images/lr_confusion_matrix.png", lr_confusion_matrix, width = 8, height = 8)
lr_confusion_matrix
```


```{r lg_model_other_metrics}
## Performance Metrics Calculation
TN <- cm$table[1,1]
FP <- cm$table[1,2]
FN <- cm$table[2,1]
TP <- cm$table[2,2]

# Calculate metrics
accuracy <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)  # Sensitivity
specificity <- TN / (TN + FP)
f1_score <- 2 * ((precision * recall) / (precision + recall))
fpr <- FP / (FP + TN)
fnr <- FN / (FN + TP)

# Create a dataframe for visualization
metrics_df <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "Specificity", "F1-Score", "FPR", "FNR"),
  Value = c(accuracy, precision, recall, specificity, f1_score, fpr, fnr)
)

# Print computed metrics 
print(metrics_df)
ratio <- metrics_df$Value[7]/metrics_df$Value[6]
```

The False Positive Rate (FPR) is higher than the False Negative Rate (FPR) by a factor of $1.25$. Considering the medical context, we try to lower the threshold to see if we can balance the ratio without losing too much in terms of accuracy. 

```{r log_model_confusion_matrix_OTHER_TRESHOLD}
## Confusion Matrix for Logistic Regression
predicted_class <- ifelse(predicted_probs >= 0.45, "Yes", "No")
cm <- confusionMatrix(as.factor(predicted_class), as.factor(actual_labels))

# Convert the confusion matrix into a table
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Prediction", "Reference", "Count")

# Plot the confusion matrix
lr_confusion_matrix <- ggplot(data = cm_table, aes(x = Reference, y = Prediction, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Confusion Matrix: Logistic Regression (threshold = 0.45)", x = "Actual", y = "Predicted")
ggsave("images/lr_confusion_matrix.png", lr_confusion_matrix, width = 8, height = 8)
lr_confusion_matrix
```

```{r lg_model_other_metrics_OTHER_TRESHOLD}
## Performance Metrics Calculation
TN <- cm$table[1,1]
FP <- cm$table[1,2]
FN <- cm$table[2,1]
TP <- cm$table[2,2]

# Calculate metrics
accuracy <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)  # Sensitivity
specificity <- TN / (TN + FP)
f1_score <- 2 * ((precision * recall) / (precision + recall))
fpr <- FP / (FP + TN)
fnr <- FN / (FN + TP)

# Create a dataframe for visualization
metrics_df <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "Specificity", "F1-Score", "FPR", "FNR"),
  Value = c(accuracy, precision, recall, specificity, f1_score, fpr, fnr)
)

# Print computed metrics
print(metrics_df)
```


```{r lr-feature-importance}
# Feature Importance
importance_lr <- as.data.frame(summary(log_model)$coefficients)
importance_lr$Feature <- rownames(importance_lr)

# Remove intercept for better visualization
importance_lr <- importance_lr[importance_lr$Feature != "(Intercept)", ]

# Rename columns for clarity
colnames(importance_lr) <- c("Estimate", "StdError", "ZValue", "PValue", "Feature")

# Plot feature importance
feature_importance <- ggplot(importance_lr, aes(x=reorder(Feature, Estimate), y=Estimate)) +
  geom_bar(stat='identity', fill='steelblue') +
  coord_flip() +
  labs(title="Coefficient Value - Logistic Regression",
       x="Feature", y="Value)") +
  theme_minimal()
ggsave("images/feature_importance.png", feature_importance, width = 8, height = 8)
feature_importance
```

We plot the absolute magnitude of every variable. The results are puzzling: cholesterol has very high importance, while pressure measurements matter much less than expected.
We hypothesize that this could be due to some sort of "correspondence" or "multicollinearity" between some variables:
- `gluc` and `cholesterol`
- `smoke` and `alco`

Let's now further investigate the relationship between some of the variables and heart disease


### Smoke & Alcohol Model
```{r}
smoke_check = glm(cardio ~ smoke + alco, data = train_data, family = binomial(link = "logit") )
summary(smoke_check)
```
Even if taken as the single predictor, it still counterintuitively contributes negatively to `cardio`. However the significance is lower than that in the complete model.
In the data exploration part, we saw that there were, numerically, minimal differences between groups with regard do this variable. 
Furthermore both of these variables are *subjective*, meaning that the answer was at the respondent's discretion. This means one could straight up lie, or even, having quit recently, answer *no* even though the effects of alcohol consumption or smoking were already present. 

### Glucose & Cholesterol Model
```{r}
gluc_check = glm(cardio ~ gluc + cholesterol, data = train_data, family = binomial(link = "logit"))
summary(gluc_check)
```
High levels of cholesterol are positively correlated with high glucose, as expected.
```{r}
interaction_gluc_cardio = glm(cardio ~ gluc*cholesterol, data = train_data, family = binomial(link = "logit"))
summary(interaction_gluc_cardio)
```
The weird effect is present just on levels of `gluc` above normal when they interact with high cholesterol. Taken individually, when they are significant, they contribute positively to the insurgence of heart diseases, as expected.  

### Pulse Pressure Model
We create a new variable, named *pulse pressure*, that combines the pressure measurements. $ap\_pu = ap\_hi - ap\_lo$.

```{r}
cardio_cl$ap_pu = cardio$ap_hi - cardio$ap_lo
log_model_ap_pu <- train(
  cardio ~ . -ap_hi-ap_lo,                    # Formula
  data = train_data,             # Dataset
  method = "glm",                # Correct method for logistic regression
  family = binomial(link = "logit"),  # Logistic regression
  trControl = ctrl,              # Cross-validation settings
  metric = "ROC"            # Appropriate metric for classification
)

print(log_model_ap_pu)
summary(log_model_ap_pu)
```



### Model diagnostics

Let's check the residual plots
```{r original_residuals 1}
fit <- glm(cardio ~ ., data = cardio, family = binomial(link = "logit"))
pred <- predict(fit, type = "response")
res <- residuals(fit, type = "response")

cardio$cardio <- factor(cardio$cardio, levels = c("No", "Yes"), labels = c(0, 1))

pearson_residuals <- (as.numeric(as.character(cardio$cardio)) - fit$fitted.values)/sqrt(fit$fitted.values*(1 - fit$fitted.values))
plot(cardio$ap_hi, pearson_residuals)
```

This graph shows *Pearson residuals*, which are the equivalent of standardized residuals for GLMs.

We can see that they are larger at the extremes. They correspond to the points we previously pointed out in the scatter plot, which represent either individuals with low blood pressure who present a heart disease, or those with high blood pressure who don't have any disease. 

```{r original_residuals 2}
binnedplot(pred,res, nclass = 1000)
```

Let's see if a polynomial regression can improve the model's fit:

```{r}
fit_poly3 <- glm(cardio ~ .+poly(ap_hi, 3), data = cardio, family = binomial(link = "logit"))

pred <- predict(fit_poly3, type = "response")
res <- residuals(fit_poly3, type = "response")
summary(fit_poly3)
pearson_residuals <- (as.numeric(as.character(cardio$cardio)) - fit$fitted.values)/sqrt(fit_poly3$fitted.values*(1 - fit_poly3$fitted.values))
plot(cardio$ap_hi, pearson_residuals, ylim = c(-10,10))

binnedplot(pred,res, nclass = 1000)
```

```{r}
fit_poly4 <- glm(cardio ~ .+poly(ap_hi, 4), data = cardio, family = binomial(link = "logit"))

pred <- predict(fit_poly4, type = "response")
res <- residuals(fit_poly4, type = "response")
summary(fit_poly4)
pearson_residuals <- (as.numeric(as.character(cardio$cardio)) - fit_poly4$fitted.values)/sqrt(fit_poly4$fitted.values*(1 - fit_poly4$fitted.values))
plot(cardio$ap_hi, pearson_residuals, ylim = c(-10,10))

binnedplot(pred,res, nclass = 1000)
br <- arm::binned.resids(pred,res)
```

It's possible to see that by considering the polynomials of the systolic blood pressure, the residuals follow less of a pattern. Further factors (like higher polynomials and multiple interactions) were also considered but they did not contribute to improving the model. 

We still get strange values for some coefficients (e.g. `gluc`).

We apply LASSO regression to see what coefficients are actually important.
```{r}
set.seed(23)
# Model matrix creation
train_data_dummies <- model.matrix(cardio ~ ., data = train_data)[, -1]  # Exclude intercept column
test_data_dummies <- model.matrix(cardio ~ ., data = test_data)[, -1]    # Exclude intercept column

X_train <- scale(train_data_dummies)
Y_train <- as.numeric(train_data$cardio) - 1  # Assuming 'cardio' is a factor variable (0/1)
X_test <- scale(test_data_dummies)
Y_test <- as.numeric(test_data$cardio) - 1  # Assuming 'cardio' is a factor variable (0/1)

# Fit the model using cv.glmnet
cvfit <- cv.glmnet(X_train, Y_train, family = "binomial", type.measure = "class", alpha = 1)
plot(cvfit)
```


```{r}
print(cvfit)
```
By using a $\lambda$ which produces a misclassification error within 1 standard error from the minimum's, we obtain similar results, while also simplifying the model: those problematic coefficients from before are shrunk to 0, underlying their secondary importance in determining the presence of heart disease. This is indeed consistent with what we had supposed in our initial exploration of the data. 

```{r}
# Get optimal lambda values
lambda_min <- cvfit$lambda.min
lambda_1se <- cvfit$lambda.1se
cat("Lambda.min:", lambda_min, "\n")
cat("Lambda.1se:", lambda_1se, "\n")
```


```{r}
# Coefficients
cat("Coefficients at lambda.min:\n")
print(coef(cvfit, s = "lambda.min"))
```


```{r}
cat("Coefficients at lambda.1se:\n")
print(coef(cvfit, s = "lambda.1se"))
```


```{r}
# Predictions on test data
predictions <- predict(cvfit, newx = X_test, s = "lambda.min", type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)
```


```{r}
# Training predictions (lambda.min)
train_predictions_min <- predict(cvfit, newx = X_train, s = "lambda.min", type = "response")
train_pred_classes_min <- ifelse(train_predictions_min > 0.5, 1, 0)
train_accuracy_min <- mean(train_pred_classes_min == Y_train)
train_roc_min <- roc(Y_train, as.vector(train_predictions_min))
train_conf_matrix_min <- table(Predicted = train_pred_classes_min, Actual = Y_train)

# Test predictions (lambda.min)
test_predictions_min <- predict(cvfit, newx = X_test, s = "lambda.min", type = "response")
test_pred_classes_min <- ifelse(test_predictions_min > 0.5, 1, 0)
test_accuracy_min <- mean(test_pred_classes_min == as.numeric(test_data$cardio) - 1)
test_roc_min <- roc(as.numeric(test_data$cardio) - 1, as.vector(test_predictions_min))
test_conf_matrix_min <- table(Predicted = test_pred_classes_min, Actual = as.numeric(test_data$cardio) - 1)

# Training predictions (lambda.1se)
train_predictions_1se <- predict(cvfit, newx = X_train, s = "lambda.1se", type = "response")
train_pred_classes_1se <- ifelse(train_predictions_1se > 0.5, 1, 0)
train_accuracy_1se <- mean(train_pred_classes_1se == Y_train)
train_roc_1se <- roc(Y_train, as.vector(train_predictions_1se))
train_conf_matrix_1se <- table(Predicted = train_pred_classes_1se, Actual = Y_train)


# Test predictions (lambda.1se)
test_predictions_1se <- predict(cvfit, newx = X_test, s = "lambda.1se", type = "response")
test_pred_classes_1se <- ifelse(test_predictions_1se > 0.5, 1, 0)
test_accuracy_1se <- mean(test_pred_classes_1se == as.numeric(test_data$cardio) - 1)
test_roc_1se <- roc(as.numeric(test_data$cardio) - 1, as.vector(test_predictions_1se))
test_conf_matrix_1se <- table(Predicted = test_pred_classes_1se, Actual = as.numeric(test_data$cardio) - 1)


# Calculate MSE for training and test sets
train_mse_min <- mean((train_predictions_min - Y_train)^2)
train_mse_1se <- mean((train_predictions_1se - Y_train)^2)
test_mse_min <- mean((test_predictions_min - (as.numeric(test_data$cardio) - 1))^2)
test_mse_1se <- mean((test_predictions_1se - (as.numeric(test_data$cardio) - 1))^2)


# Create a summary table
results <- data.frame(
  Metric = c(
    "Training Accuracy", "Training AUC", "Training MSE",
    "Test Accuracy", "Test AUC", "Test MSE"
  ),
  Lambda_Min = c(
    train_accuracy_min, auc(train_roc_min), train_mse_min,
    test_accuracy_min, auc(test_roc_min), test_mse_min
  ),
  Lambda_1SE = c(
    train_accuracy_1se, auc(train_roc_1se), train_mse_1se,
    test_accuracy_1se, auc(test_roc_1se), test_mse_1se
  )
)

print("Summary of Results:")
results
```

```{r}
fits1 <- glmnet(X_train, Y_train, family = "binomial", type.measure = "class", alpha = 1, lambda = lambda_1se)

# Get predicted probabilities
predicted_probs <- predict(fits1, newx = X_test, type = "response")
predicted_class <- ifelse(predicted_probs >= 0.5, "Yes", "No")
actual_labels <- as.factor(Y_test)
levels(actual_labels) <- c("No", "Yes")

# Create confusion matrix
conf_matrix <- confusionMatrix(
  factor(predicted_class),
  factor(actual_labels),
  positive = "Yes"
)
lasso_sensitivity <- conf_matrix$byClass["Sensitivity"]
lasso_specificity <- conf_matrix$byClass["Specificity"]

# Manually compute the area under ROC curve
roc_curve_lasso <- roc(actual_labels, as.numeric(predictions))

# ROC curve
plot(roc_curve_lasso, main = "ROC Curve for LASSO", col = "blue", lwd = 2)
```
```{r}
roc_curve_lasso$auc
```

Lastly, let's fit the polynomial models this way as well

### Polynomial of degree 3
```{r}
# Cross-validation
poly3 <- train(
  cardio ~ . + poly(ap_hi, 3),   # Formula
  data = train_data,             # Dataset
  method = "glm",                # Correct method for logistic regression
  family = binomial(link = "logit"),  # Logistic regression
  trControl = ctrl,              # Cross-validation settings
  metric = "ROC"            # Appropriate metric for classification
)

# General results
poly3$results

# Get predicted probabilities
predictions <- predict(poly3, newdata = test_data, type = "prob")
predicted_probs <- predictions[, "Yes"] 
actual_labels <- test_data$cardio 

# Manually compute the area under ROC curve
roc_curve_poly3 <- roc(actual_labels, predicted_probs)

# ROC curve
plot(roc_curve_poly3, main = "ROC Curve for poly3", col = "blue", lwd = 2)
```
```{r}
roc_curve_poly3$auc
```

### Polynomial of degree 4
```{r}

# Cross-validation
poly4 <- train(
  cardio ~ . + poly(ap_hi, 4),   # Formula
  data = train_data,             # Dataset
  method = "glm",                # Correct method for logistic regression
  family = binomial(link = "logit"),  # Logistic regression
  trControl = ctrl,              # Cross-validation settings
  metric = "ROC"            # Appropriate metric for classification
)

# General results
poly4$results

# Get predicted probabilities
predictions <- predict(poly4, newdata = test_data, type = "prob")
predicted_probs <- predictions[, "Yes"] 
actual_labels <- test_data$cardio 

# Manually compute the area under ROC curve
roc_curve_poly4 <- roc(actual_labels, predicted_probs)

# ROC curve
plot(roc_curve_poly4, main = "ROC Curve for poly4", col = "blue", lwd = 2)
```
```{r}
roc_curve_poly4$auc
```


### Splines

```{r spline}
set.seed(23)

library(splines)
library(MASS)
library(caret)
library(pROC)

fit <- glm(cardio ~ ns(age, df = 4) + ns(height, df = 4) + ns(weight, df = 4) + ns(ap_hi, df = 4) +
              ns(ap_lo, df = 4) + gender + cholesterol + gluc + smoke + alco + active, 
            data = train_data, family = binomial(link = "logit"))

fit_step <- stepAIC(fit, direction = "both", trace = FALSE)

summary(fit_step)

predicted_probs <- predict(fit_step, newdata = test_data, type = "response")

predicted_class <- ifelse(predicted_probs >= 0.40, "Yes", "No")

actual_labels <- test_data$cardio
cm <- confusionMatrix(as.factor(predicted_class), as.factor(actual_labels))

cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Prediction", "Reference", "Count")

lr_confusion_matrix <- ggplot(data = cm_table, aes(x = Reference, y = Prediction, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Confusion Matrix: Logistic Regression with Splines (threshold = 0.4)", x = "Actual", y = "Predicted")
ggsave("images/lr_confusion_matrix.png", lr_confusion_matrix, width = 8, height = 8)
lr_confusion_matrix

roc_curve_splines <- roc(actual_labels, predicted_probs)
plot(roc_curve_splines, main = "ROC Curve", col = "blue", lwd = 2)
auc_spline <- auc(roc_curve_splines)
print(auc_spline)

png("images/roc_curve_splines.png", width = 800, height = 600)
plot(roc_curve_splines, main = "ROC Curve", col = "blue", lwd = 2)
dev.off()

TN <- cm$table[1, 1]
FP <- cm$table[1, 2]
FN <- cm$table[2, 1]
TP <- cm$table[2, 2]

accuracy <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
specificity <- TN / (TN + FP)
f1_score <- 2 * ((precision * recall) / (precision + recall))
fpr <- FP / (FP + TN)
fnr <- FN / (FN + TP)
fnr_fpr_ratio <- fnr / fpr

splines_metrics_df <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "Specificity", "F1-Score", "FPR", "FNR", "FNR/FPR Ratio"),
  Value = c(accuracy, precision, recall, specificity, f1_score, fpr, fnr, fnr_fpr_ratio)
)

print(splines_metrics_df)
```


```{r spline decision boundary}
get_mode <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]  
}
ap_hi_range <- seq(min(train_data$ap_hi), max(train_data$ap_hi), length.out = 100)
ap_lo_range <- seq(min(train_data$ap_lo), max(train_data$ap_lo), length.out = 100)
grid <- expand.grid(ap_hi = ap_hi_range, ap_lo = ap_lo_range)
grid$age <- median(train_data$age, na.rm = TRUE)
grid$height <- median(train_data$height, na.rm = TRUE)
grid$weight <- median(train_data$weight, na.rm = TRUE)
grid$cholesterol <- get_mode(train_data$cholesterol)
grid$gluc <- get_mode(train_data$gluc)
grid$smoke <- get_mode(train_data$smoke)
grid$alco <- get_mode(train_data$alco)
grid$active <- get_mode(train_data$active)
grid$gender <- get_mode(train_data$gender)
ap_hi_spline <- ns(grid$ap_hi, df = 4)
ap_lo_spline <- ns(grid$ap_lo, df = 4)
grid_splines <- cbind(ap_hi_spline, ap_lo_spline)
grid_full <- cbind(grid, grid_splines)
colnames(grid_full) <- c("ap_hi", "ap_lo", "age", "height", "weight", "cholesterol", "gluc", "smoke", "alco", "active", "gender", paste0("ns_ap_hi", 1:4), paste0("ns_ap_lo", 1:4))
grid_full$gender <- factor(grid_full$gender, levels = levels(train_data$gender))
grid_full$cholesterol <- factor(grid_full$cholesterol, levels = levels(train_data$cholesterol))
grid_full$gluc <- factor(grid_full$gluc, levels = levels(train_data$gluc))
grid_full$smoke <- factor(grid_full$smoke, levels = levels(train_data$smoke))
grid_full$alco <- factor(grid_full$alco, levels = levels(train_data$alco))
grid_full$active <- factor(grid_full$active, levels = levels(train_data$active))

grid_full$predicted_probs <- predict(fit_step, newdata = grid_full, type = "response")
grid_full$predicted_class <- ifelse(grid_full$predicted_probs >= 0.5, "Yes", "No")

# Add new data for "well above normal" cholesterol
grid_well_above <- grid_full
grid_well_above$cholesterol <- "well above normal"
grid_well_above$predicted_probs <- predict(fit_step, newdata = grid_well_above, type = "response")
grid_well_above$predicted_class <- ifelse(grid_well_above$predicted_probs >= 0.5, "Yes", "No")

# Add new data for "above normal" cholesterol
grid_above_normal <- grid_full
grid_above_normal$cholesterol <- "above normal"
grid_above_normal$predicted_probs <- predict(fit_step, newdata = grid_above_normal, type = "response")
grid_above_normal$predicted_class <- ifelse(grid_above_normal$predicted_probs >= 0.5, "Yes", "No")

# Add new data for "normal" cholesterol (this is the default scenario)
grid_normal <- grid_full
grid_normal$cholesterol <- "normal"
grid_normal$predicted_probs <- predict(fit_step, newdata = grid_normal, type = "response")
grid_normal$predicted_class <- ifelse(grid_normal$predicted_probs >= 0.5, "Yes", "No")

ggplot(grid_full, aes(x = ap_hi, y = ap_lo, color = predicted_class)) +
  geom_tile(aes(fill = predicted_class), alpha = 0.3) + 
  geom_point(data = train_data, aes(x = ap_hi, y = ap_lo, color = factor(cardio)), linewidth = 2) + 
  scale_color_manual(values = c("blue", "red")) +  
  scale_fill_manual(values = c("blue", "red")) +
  geom_contour(data = grid_well_above, aes(z = predicted_probs, linetype = "Well Above Normal"), 
               color = "black", size = 1, breaks = 0.5) +
  geom_contour(data = grid_above_normal, aes(z = predicted_probs, linetype = "Above Normal"), 
               color = "gray25", size = 1, breaks = 0.5) +
  geom_contour(data = grid_normal, aes(z = predicted_probs, linetype = "Normal"), 
               color = "gray95", size = 1, breaks = 0.5) +
  labs(title = "Decision Boundary for Logistic Regression Model with Cholesterol Levels",
       x = "ap_hi", y = "ap_lo", color = "Cardio Outcome", linetype = "Cholesterol Level") +
  scale_linetype_manual(values = c("Normal" = "solid", "Above Normal" = "solid", "Well Above Normal" = "solid")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.box = "horizontal",
        legend.title = element_text(size = 8),  # Even smaller legend title size
        legend.text = element_text(size = 6),   # Even smaller legend text size
        legend.key.size = unit(0.6, "lines"),   # Reduce size of legend keys
        legend.key.width = unit(0.8, "lines"),  # Reduce the width of the legend keys
        legend.spacing.x = unit(0.5, "mm"),     # Reduce space between legend items
        legend.spacing.y = unit(0.5, "mm")) +   # Reduce vertical space between items
  guides(linetype = guide_legend(ncol = 2))  # Split the legend into two columns
```



## Results

Let's compare the results:
```{r}
# Plot the ROC curves
plot(roc_curve_rf, col = "red", lwd = 2, main = "ROC Curves for the main models")
plot(roc_curve_lr, col = "blue", lwd = 2, add = TRUE)
plot(roc_curve_lasso, col = "pink", lwd = 2, add = TRUE)
plot(roc_curve_poly3, col = "green", lwd = 2, add = TRUE)
plot(roc_curve_poly4, col = "orange", lwd = 2, add = TRUE)
plot(roc_curve_splines, col = "purple", lwd = 2, add = TRUE)

# Add a legend
legend("bottomright", legend = c("Random Forest", "Basic Logistic Regression", "LASSO", "Polynomial Regression 3",  "Polynomial Regression 4", "Splines"), col = c("red", "blue", "pink", "green", "orange", "purple"), lwd = 2)

# Save image
png("images/roc_results.png", width = 800, height = 600)
plot(roc_curve_rf, col = "red", lwd = 2, main = "ROC Curves for the main models")
plot(roc_curve_lr, col = "blue", lwd = 2, add = TRUE)
plot(roc_curve_lasso, col = "pink", lwd = 2, add = TRUE)
plot(roc_curve_poly3, col = "green", lwd = 2, add = TRUE)
plot(roc_curve_poly4, col = "orange", lwd = 2, add = TRUE)
plot(roc_curve_splines, col = "purple", lwd = 2, add = TRUE)
legend("bottomright", legend = c("Random Forest", "Basic Logistic Regression", "LASSO", "Polynomial Regression 3",  "Polynomial Regression 4", "Splines"), col = c("red", "blue", "pink", "green", "orange", "purple"), lwd = 2)
dev.off()

# Model results
results <- data.frame(
  Model = c("Random Forest", "Basic Logistic Regression", "LASSO", "Polynomial Regression 3", "Polynomial Regression 4", "Splines"),
  AUC = c(
    rf_model$results$ROC,
    log_model$results$ROC,
    roc_curve_lasso$auc,
    poly3$results$ROC,
    poly4$results$ROC,
    auc_spline
  ),
  Sensitivity = c(
    rf_model$results$Sens,
    log_model$results$Sens,
    lasso_sensitivity,
    poly3$results$Sens,
    poly4$results$Sens,
    splines_metrics_df$Value[splines_metrics_df$Metric == "Recall (Sensitivity)"]
  ),
  Specificity = c(
    rf_model$results$Spec,
    log_model$results$Spec,
    lasso_specificity,
    poly3$results$Spec,
    poly4$results$Spec,
    splines_metrics_df$Value[splines_metrics_df$Metric == "Specificity"]
  )
)
print(results)
```

As it is shown above, the four main models get similar results in the task of classifying the data. Despite its slightly better performance, the lack of explainability makes the random forest a less suitable model for this context, leaving the logistic regressions as the best choices. 


## References

