---
title: "Cardiovascular Disease Data Analysis"
author: Giovanni Billo, Muhammad Mubashar Shahzad, Carlos Velázquez Fernández, Fabio Vicig
date: "2025-01-07"
output: 
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
references:
  - id: narloch1995influence
    title: "Influence of breathing technique on arterial blood pressure during heavy weight lifting"
    author:
      - family: "Narloch"
        given: "Joseph A"
      - family: "Brandstater"
        given: "Murray E"
    container-title: "Archives of physical medicine and rehabilitation"
    volume: 76
    issue: 5
    page: "457-462"
    issued:
      year: 1995
    publisher: "Elsevier"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(corrplot)
library(reshape2)
library(ggplot2)
library(dplyr)
library(patchwork)
library(gridExtra)
library(dendextend)
# Assess clusters
library(cluster)
```

## Introduction

The analysis is centered around the prediction of presence of cardiovascular disease, given data about patients. [TABLE OF CONTENTS, SUMMARY OF WHAT HAS BEEN DONE]

## Data exploration

Firstly, an exploratory analysis has been conducted, with the purpose of identifying problems and have at a glance an idea of meaning and distribution of the variables.

```{r 'Explorative analysis and data cleaning'}
cardio_with_id <- read.csv("data/cardio_train.csv", sep= ";", header=TRUE)
cardio <- cardio_with_id[,-1]
sum(as.numeric(is.na(cardio)))
str(cardio)
summary(cardio)
par(mfrow = c(2,2))
```

The dataset contains 70000 observations of 12 variables. The source doesn't specify how the data was recovered or generated, and it has not been possible to find it elsewhere than Kaggle. The variables are the following:\

- **cardio** : objective variable of our analysis. \
- **age** : age of the patient, in days.\
- **gender** : categorical variable which is 1 for females and 2 for males.\
- **height** : height of the patient, in cm.\
- **weight** : weight of the patient, in kg.\
- **ap_hi** : systolic blood pressure.\
- **ap_lo** : diastolic blood pressure.\
- **cholesterol** : categorical variable with three levels of cholesterol: 1: normal, 2: above normal, 3: well above normal.\
- **gluc** : categorical variable with three levels of glucose: 1: normal, 2: above normal, 3: well above normal.\
- **smoke** : binary variable, representing smokers.\
- **alco** : binary variable, representing alcohol intake.\
- **active** : binary variable, representing physical activity.

The last three variables are marked as "subjective", meaning that this info is supplied by patients themselves (probably through surveying) and therefore, being based on the patient's discretion, is not necessarily true,  \
\
There are no NA values. However, from the summary we can see that there are variables with impossible values: specifically, *ap_hi* and *ap_lo* have in some cases negative values, which don't make sense even for a dead person, and in other cases extremely high values.
The highest pressure recorded in an individual was 370/360 [@narloch1995influence].

```{r}
summary(cardio[(cardio$ap_hi<0|cardio$ap_hi>400|cardio$ap_lo<0|cardio$ap_lo>400|cardio$ap_hi<cardio$ap_lo),])
cardio <- cardio[!(cardio$ap_hi < 0 | cardio$ap_hi > 400 | cardio$ap_lo < 0 | cardio$ap_lo > 400 | cardio$ap_hi < cardio$ap_lo),]

# save a copy of cardio with filtered values but without factor columns for clustering later on
cardio_cl = cardio
```

1274 observations have completely unrealistic values for either systolic or diastolic pressure (or both). In 274 of them, the systolic pressure is less than the diastolic, which is also not possible. These wrong metrics are less than 2% of our dataset, so the most straightforward approach is just to eliminate the problematic rows. \

```{r}
cardio$gender <- factor(cardio$gender)
levels(cardio$gender) = c("F", "M")
head(cardio$gender)

cardio$cardio <- as.factor(cardio$cardio)
head(cardio$cardio)

cardio$cholesterol <- factor(cardio$cholesterol)
levels(cardio$cholesterol) = c("normal", "above normal", "well above normal")
head(cardio$cholesterol)

cardio$gluc <- factor(cardio$gluc)
levels(cardio$gluc) = c("normal", "above normal", "well above normal")
head(cardio$gluc)

cardio$smoke <- factor(cardio$smoke)
levels(cardio$smoke) = c("No", "Yes")
head(cardio$smoke)

cardio$alco <- factor(cardio$alco)
levels(cardio$alco) = c("No", "Yes")
head(cardio$alco)

cardio$active <- factor(cardio$active)
levels(cardio$active) = c("No", "Yes")
head(cardio$active)

```

```{r age_categorizing}
cardio$age = ceiling(cardio$age/365)
cardio$age_cat <- ifelse(cardio$age >= 18 & cardio$age <= 35, 'Y', 
                         ifelse(cardio$age > 35 & cardio$age <= 55, 'S', 'E'))
table(cardio$age_cat)
```
There are virtually no observations concerning young people (< 35 years old): we remove them for clarity and we focus on senior and elderly.
```{r}
cardio = cardio %>% filter(age_cat == "S" | age_cat == "E")
cardio$age_cat <- factor(cardio$age_cat)
levels(cardio$age_cat) = c("Senior", "Elderly")
head(cardio$age_cat)
```



```{r}
table(cardio$cardio)
```
The dataset is balanced with respect to the target variable.

Let's divide our dataset between numerical and categorical variables and analyze what are, if any, the relationships between the variables.
```{r eval=FALSE, include=FALSE}
numerical <- cardio[, c("age", "height", "weight", "ap_hi", "ap_lo")]
categorical <- cardio[!(names(cardio) %in% c("age", "height", "weight", "ap_hi", "ap_lo"))]
```

```{r eval=FALSE, include=FALSE}
plot.new()

pairs(numerical,
      labels = colnames(numerical),  
      pch = 21,
      bg = rainbow(2)[(cardio$cardio)],
      col = rainbow(2)[(cardio$cardio)], 
      main = "Cardiovascular disease dataset (numerical variables)",
      row1attop = TRUE,         # If FALSE, changes the direction of the diagonal
      gap = 1,                  # Distance between subplots
      cex.labels = NULL,        # Size of the diagonal text
      font.labels = 1)
# Add the legend
legend(
  "topright",                           # Position of the legend
  legend = c("No disease", "Disease"), # Labels for the categories
  col = rainbow(2),                     # Colors corresponding to categories
  pch = 21,                             # Point shape used in the plot
  pt.bg = rainbow(2),                   # Background color for the points
  bty = "n"                             # No border around the legend box
)
```
Let's visualize the distribution of the numerical variables as well
```{r eval=FALSE}
# Define the number of variables
num_vars <- ncol(numerical)

# Set up a grid layout for larger histograms
par(mfrow = c(ceiling(sqrt(num_vars)), ceiling(sqrt(num_vars)))) # Adjust grid based on number of variables
par(mar = c(4, 4, 2, 1)) # Adjust margins for better spacing

# Loop through each variable and create histograms
for (i in 1:num_vars) {
    hist(
        numerical[[i]], 
        col = rgb(0, 1, 1, alpha = 0.5), 
        main = colnames(numerical)[i], 
        xlab = "Value", 
        ylab = "Frequency",
        border = "white"
    )
}

```

Correlation analysis
```{r}
cor_cardio = round(cor(as.data.frame(numerical)), 2)
melt_cardio = melt(cor_cardio)

ggplot(data = melt_cardio, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  geom_text(aes(Var2, Var1, label = value), size = 5) +
  scale_fill_gradient2(low = "blue", high = "red",
                       limit = c(-1,1), name="Correlation") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.background = element_blank())
```
The heatmap confirms some (perhaps rather intuitive) relationships:
- Diastolic and systolic blood pressure are heavily correlated
- Both weight and age are positively correlated with higher diastolic and systolic blood pressure, the former slightly more than the latter.

Let's now explore the possible relationships of the other categorical variables with cardio.

```{r}
summary_table = as.data.frame(table(cardio$age_cat, cardio$cardio))
colnames(summary_table) <- c("age_cat", "cardio", "count")

ggplot(summary_table, aes(x = age_cat, y = count, fill = as.factor(cardio))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Counts of Cardio Diseases by Age Category",
    x = "Age Category",
    y = "Count",
    fill = "Cardio"
  ) +
  theme_minimal()
```
We see that, interestingly, a higher occurrence of heart disease occurs in senior patients (from 35 to 55) than in elderly patients: actually the majority of the first category is affected by heart disease, while rather surprisingly the majority of the elderly are not.

We use the Chi-squared test to see to which degree the variables are correlated with cardio.
```{r}
categorical_vars = c("gender", "cholesterol", "gluc", "smoke", "alco", "active", "age_cat")
p_values = c()
for (i in categorical_vars){
  corr = chisq.test(table(cardio$cardio, cardio[, i]))
  p_values[i] = corr$p.value
}
which(p_values < 0.05)
```
We see that all of the variables available, beside gender, seem to be very relevant in determining the insurgence of heart diseases.

Let's now test the mean differences between groups for dystolic and systolic blood pressure.
```{r}
continuous_vars = c("ap_hi", "ap_lo")
mean_diff_results <- data.frame(
    Variable = character(),
    Group = character(),
    Test = character(),
    P_Value = numeric(),
    stringsAsFactors = FALSE
)
# t_test = t.test(cardio[, "ap_hi"] ~ cardio[, "smoke"])
# print(t_test)
# 
# anova_test = aov(cardio[, "ap_hi"] ~ cardio[, "gluc"])

for (cont_var in continuous_vars) {
  for (cat_var in categorical_vars) {
    num_levels = length(unique(cardio[, cat_var]))
    if (num_levels == 2) {
      # Perform t-test
      t_test = t.test(cardio[, cont_var] ~ cardio[, cat_var])
      p_value = t_test$p.value
      test_type = "t-test"
    } else {
      # Perform ANOVA
      anova_test = aov(cardio[, cont_var] ~ cardio[, cat_var])
      p_value = summary(anova_test)[[1]]["Pr(>F)"][1]
      test_type = "ANOVA"
    }
    # Append results
    results = data.frame(
                                Variable = cont_var, 
                                Group = cat_var, 
                                Test = test_type,
                                P_Value = p_value
                              )
    mean_diff_results = bind_rows(mean_diff_results, results)
  }
}
# !!! IDK what is wrong with this loop

# View results
print(mean_diff_results)
```
Once again, there definitely are some significant differences between the groups considered.
```{r}
# just like in the example, we try to do "visual" clustering, see if there are any evident clusters by plotting data
p1 <- ggplot(cardio, aes(cardio, ap_hi)) + geom_point(aes(color = cholesterol))
p2 <- ggplot(cardio, aes(cardio, ap_hi)) + geom_point(aes(color = age_cat))
p3 <- ggplot(cardio, aes(cardio, ap_hi)) + geom_point(aes(color = smoke))
p4 <- ggplot(cardio, aes(cardio, ap_hi)) + geom_point(aes(color = active))
p5 <- ggplot(cardio, aes(cardio, ap_hi)) + geom_point(aes(color = alco))
p6 <- ggplot(cardio, aes(cardio, ap_hi)) + geom_point(aes(color = gluc))
```

```{r}
grid.arrange(p1, p2, p3, p4)
```

```{r}
grid.arrange(p5, p6, p7)
```
```{r}
p8 <- ggplot(cardio, aes(cardio, ap_lo)) + geom_point(aes(color = cholesterol))
p9 <- ggplot(cardio, aes(cardio, ap_lo)) + geom_point(aes(color = age_cat))
p10 <- ggplot(cardio, aes(cardio, ap_lo)) + geom_point(aes(color = smoke))
p11 <- ggplot(cardio, aes(cardio, ap_lo)) + geom_point(aes(color = active))
p12 <- ggplot(cardio, aes(cardio, ap_lo)) + geom_point(aes(color = alco))
p13 <- ggplot(cardio, aes(cardio, ap_lo)) + geom_point(aes(color = gluc))

grid.arrange(p8, p9, p10, p11)
```

```{r}
grid.arrange(p12, p13)
```
While no group shows a clean division with respect to `cardio`, values of both `gluc` and `cholesterol` above normal or well above normal seem to be particularly frequent along with the occurrence of heart diseases.
```{r}
count_disease_with_gluc = count(cardio %>% filter(gluc != "normal" & cardio == 1))$n
count_disease_with_chol = cardio %>% filter(cholesterol != "normal" && cardio == 1)$n # opr maybe not: apparently there are no occurrences of disease with cholesterol above normal
```

## Clustering
### K-means clustering
```{r}
# Scale the data to normalize the values
cardio_data_scaled <- scale(cardio_cl)

# View the first few rows of the dataset
head(cardio_data_scaled)

# Perform K-means clustering
set.seed(123)  
silhouette_scores = c()

# Compute the distance matrix for the scaled data
sample_cardio_data_scaled = sample(cardio_data_scaled, 10000)
dist_matrix <- dist(sample_cardio_data_scaled)

# Try out different numbers of clustering and see which one has the highest silhouette score (meaning it best divides the data)
# Due to computational constraints we select just a subset of the original observations, 1/7, to tune this hyperparameter
for (n_clusters in 2:10) {
  # Perform k-means clustering
  kmeans_result <- kmeans(sample_cardio_data_scaled, centers = n_clusters)
  
  # Extract cluster labels from the k-means result
  cluster_labels <- kmeans_result$cluster
  
  # Calculate silhouette values
  sil <- silhouette(cluster_labels, dist_matrix)
  
  # Store the average silhouette score for the current number of clusters
  silhouette_scores[n_clusters] <- mean(sil[, 3])
}
max_silhouette_kmeans = which.max(silhouette_scores)

best_kmeans_result <- kmeans(cardio_data_scaled, centers = max_silhouette_kmeans)


# Add the cluster assignment to the original data
cardio$Cluster <- as.factor(best_kmeans_result$cluster)

# Visualize the clusters across different dimensions
cl_p1 = ggplot(cardio, aes(cardio, ap_hi, color = Cluster)) +
  geom_point() +
  labs(title = "K-Means Clustering (cardio vs ap_hi)")
cl_p2 = ggplot(cardio, aes(cardio, ap_lo, color = Cluster)) +
  geom_point() +
  labs(title = "K-Means Clustering (cardio vs ap_lo)")
cl_p3 = ggplot(cardio, aes(cardio, weight, color = Cluster)) +
  geom_point() +
  labs(title = "K-Means Clustering (cardio vs ap_lo)")


grid.arrange(cl_p1, cl_p2)
cl_p3 # the difference is much more blurried here
```
Plotting along the cardio and both types of pressures, the clusters seem to neatly separate the data along some pressure values. 

Let's test for mean differences across them

### Hierarchical clustering
```{r}
# Calculate the distance matrix
dist_matrix <- dist(sample(cardio_data_scaled, 10000))

# Perform hierarchical clustering
hclust_result <- hclust(dist_matrix, method = "ward.D2")

silhouette_scores = c()
# Compute silhouette
for (n_clusters in 2:10){
  cluster_labels <- cutree(hclust_result, k = n_clusters)
  sil <- silhouette(cluster_labels, dist_matrix)
  silhouette_scores[n_clusters] = mean(sil[, 3])
}

best_n_cluster = which.max(silhouette_scores)

# visualize the best performing cluster
dend <- as.dendrogram(hclust_result)

# Color branches by clusters
dend <- color_branches(dend, k = 3)

# Plot the dendrogram
plot(dend, main = "Hierarchical Clustering with Colored Clusters")
```

```{r}
# Plot silhouette (!!! IDK what is wrong with this as well)
best_cluster_labels <- cutree(hclust_result, k = best_n_cluster)
best_sil <- silhouette(best_cluster_labels, dist_matrix)
plot(best_sil, main = "Best-performing Silhouette Plot for Clustering")
```

We now augment our dataset with the cluster group. 
```{r}
cardio_data_scaled$cluster <- best_cluster_labels

# View the dataset with cluster labels
head(cardio_data_scaled)

```

## References
